---
title: "Scoring a set of genes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring a set of genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

`CcSinglecell` grew out of a colorectal-cancer study that ranked genes by
combining three signals — expression variability (MAD), switch-like dynamics
along a pseudotime trajectory (switchDE), and cancer-miRNA targeting — and then
tested how well the top genes stratified patient survival. This vignette shows
how to score an arbitrary set of genes and visualize the result, either from R
or from the command line.

```{r load}
library(CcSinglecell)
```

## Example data

The package ships small, synthetic example files under `inst/extdata/`:

```{r data}
ext <- function(f) system.file("extdata", f, package = "CcSinglecell")

expr  <- as.matrix(read.csv(ext("example_expression.csv"), row.names = 1, check.names = FALSE))
mirna <- as.matrix(read.csv(ext("example_mirna.csv"), row.names = 1, check.names = FALSE))
genes <- read_gene_list(ext("example_genes.csv"))

dim(expr)
head(genes)
```

## Score a gene set

`score_gene_set()` computes each available metric genome-wide, restricts to the
genes you asked about, and blends them into a single `combined` score. With only
an expression matrix you get MAD scoring; add a miRNA matrix (and/or a pseudotime
vector) to bring in the other metrics.

```{r score}
scores <- score_gene_set(genes, expr, mirna_matrix = mirna)
head(scores)
attr(scores, "weights")
```

## Visualize the scores

```{r plot-scores}
plot_gene_scores(scores, top_n = 15)
```

The stacked contribution plot shows how much each metric drives each gene's
combined score:

```{r plot-contrib}
plot_score_contributions(scores, top_n = 15)
```

## From the command line

The same workflow is available as a `ccscore` executable installed with the
package:

```sh
ccscore score \
  --genes      example_genes.csv \
  --expression example_expression.csv \
  --mirna      example_mirna.csv \
  --out        results/
```

This writes `results/scores.csv`, `results/gene_scores.png`, and
`results/score_contributions.png`. Find the installed script with:

```{r cli-path, eval = FALSE}
system.file("bin", "ccscore", package = "CcSinglecell")
```

## Optional: survival stratification

If you also have a survival data frame (patients in rows; gene-expression
columns plus `days_to_last_follow_up` and `vital_status`), you can fit a
penalized Cox model on the gene set and draw a Kaplan-Meier curve. This mirrors
the paper's primary result.

```{r survival, eval = requireNamespace("survminer", quietly = TRUE) && requireNamespace("glmnet", quietly = TRUE)}
cox_df <- read.csv(ext("example_survival.csv"), row.names = 1, check.names = FALSE)
model  <- fit_cox_model(cox_df, gene_names = genes)
risk   <- calculate_risk_scores(cox_df, model$active_genes, model$coefficients)
plot_km_curve(risk, title = "Example cohort")
```

The equivalent command is:

```sh
ccscore survival --genes example_genes.csv --survival example_survival.csv --out results/
```
