---
title: "Scoring a set of genes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scoring a set of genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

`CcSinglecell` grew out of a colorectal-cancer study that ranked genes by
combining three signals — expression variability (MAD), switch-like dynamics
along a pseudotime trajectory (switchDE), and cancer-miRNA targeting — and then
tested how well the top genes stratified patient survival. This vignette shows
how to score an arbitrary set of genes and visualize the result, either from R
or from the command line.

```{r load}
library(CcSinglecell)
```

## Example data

The package ships small, synthetic example files under `inst/extdata/`:

```{r data}
ext <- function(f) system.file("extdata", f, package = "CcSinglecell")

expr  <- as.matrix(read.csv(ext("example_expression.csv"), row.names = 1, check.names = FALSE))
mirna <- as.matrix(read.csv(ext("example_mirna.csv"), row.names = 1, check.names = FALSE))
genes <- read_gene_list(ext("example_genes.csv"))

dim(expr)
head(genes)
```

## Score a gene set

`score_gene_set()` computes each available metric genome-wide, restricts to the
genes you asked about, and blends them into a single `combined` score. With only
an expression matrix you get MAD scoring; add a miRNA matrix (and/or a pseudotime
vector) to bring in the other metrics.

```{r score}
scores <- score_gene_set(genes, expr, mirna_matrix = mirna)
head(scores)
attr(scores, "weights")
```

## Visualize the scores

```{r plot-scores}
plot_gene_scores(scores, top_n = 15)
```

The stacked contribution plot shows how much each metric drives each gene's
combined score:

```{r plot-contrib}
plot_score_contributions(scores, top_n = 15)
```

## Scoring across arbitrary omics layers

`score_gene_set()` handles the built-in expression / pseudotime / miRNA
metrics. To integrate *any* number of omics layers — including your own scoring
functions — use `score_multiomics()` with one `omics_block()` per layer. Each
block pairs a data object with a metric, which is either a built-in name
(`"mad"`, `"mirna"`, `"switchde"`) or a function mapping the data to a named
gene ranking.

```{r multiomics}
# Treat the example expression matrix as a second "pseudo-omics" layer scored by
# a custom metric (per-gene standard deviation) to illustrate the interface.
blocks <- list(
  omics_block("variability", expr, "mad"),
  omics_block("mirna", mirna, "mirna"),
  omics_block("dispersion", expr, function(m) sort(apply(m, 1, sd), decreasing = TRUE))
)
mo <- score_multiomics(genes, blocks)
head(mo)
```

The combining machinery (`combine_rankings()`, `optimize_weights()`) is no
longer capped at three metrics, so methylation, CNV, or proteomics layers plug
in the same way. See `ROADMAP.md` for the planned learned-integration and
regulatory-activity backends.

### Expression-anchored miRNA activity

The bundled `example_mirna.csv` is a *predicted* interaction map (how many
cancer miRNAs are annotated to target each gene). `calculate_mirna_activity()`
goes further: given matched miRNA and gene expression, it scores genes by the
*observed* anti-correlation with their targeting miRNAs — repression you can
actually see in the samples, rather than annotation alone.

```{r mirna-activity}
mirna_expr <- as.matrix(read.csv(ext("example_mirna_expression.csv"),
  row.names = 1, check.names = FALSE))
target <- as.matrix(read.csv(ext("example_mirna.csv"),
  row.names = 1, check.names = FALSE))
gene_expr <- as.matrix(read.csv(ext("example_expression.csv"),
  row.names = 1, check.names = FALSE))

activity <- calculate_mirna_activity(mirna_expr, gene_expr, target)
head(activity)
```

Drop it into a multi-omics run as its own block:

```{r mirna-activity-block}
block <- omics_block(
  "mirna_activity",
  list(mirna_expr = mirna_expr, gene_expr = gene_expr, target_matrix = target),
  "mirna_activity"
)
head(score_multiomics(genes, list(omics_block("variability", gene_expr, "mad"), block)))
```

## From the command line

The same workflow is available as a `ccscore` executable installed with the
package:

```sh
ccscore score \
  --genes      example_genes.csv \
  --expression example_expression.csv \
  --mirna      example_mirna.csv \
  --out        results/
```

This writes `results/scores.csv`, `results/gene_scores.png`, and
`results/score_contributions.png`. Find the installed script with:

```{r cli-path, eval = FALSE}
system.file("bin", "ccscore", package = "CcSinglecell")
```

## Optional: survival stratification

If you also have a survival data frame (patients in rows; gene-expression
columns plus `days_to_last_follow_up` and `vital_status`), you can fit a
penalized Cox model on the gene set and draw a Kaplan-Meier curve. This mirrors
the paper's primary result.

```{r survival, eval = requireNamespace("survminer", quietly = TRUE) && requireNamespace("glmnet", quietly = TRUE)}
cox_df <- read.csv(ext("example_survival.csv"), row.names = 1, check.names = FALSE)
model  <- fit_cox_model(cox_df, gene_names = genes)
risk   <- calculate_risk_scores(cox_df, model$active_genes, model$coefficients)
plot_km_curve(risk, title = "Example cohort")
```

The equivalent command is:

```sh
ccscore survival --genes example_genes.csv --survival example_survival.csv --out results/
```
